---
title: "Group comparison example"
author: "Guillaume A. Rousselet & Rand R. Wilcox"
date: "`r Sys.Date()`"
output:
  # pdf_document:
  #   fig_caption: no
  #   number_sections: no
  #   toc: yes
  #   toc_depth: 2
    github_document:
    html_preview: yes
    toc: yes
    toc_depth: 2
---

# Dependencies

```{r}
library(tibble)
library(ggplot2)
source("./code/functions.R")
source("./code/theme_gar.txt")
```

# Load data

Data from Clark et al. (2012).
```{r}
A1 <- read.table(file = "./data/A1_dat.txt")
a <- fac2list(A1$CESD,A1$edugp)
```

# Illustrate results

Horizontal black = means. 
Horizontal grey = medians.  

```{r}
df <- tibble(gp = factor(c(rep("High school", length(a[[2]])), rep("Advanced degree", length(a[[5]])))),
             score = c(a[[2]],a[[5]]))

ggplot(df, aes(x=gp, y=score)) + theme_gar +
  geom_jitter(shape = 21, width = 0.1, fill = "grey") +
  stat_summary(fun = "mean", geom = "crossbar", width = 0.5, size = 0.2) +
  stat_summary(fun = "median", geom = "crossbar", width = 0.5, size = 0.2, colour = "grey") +
  labs(x = "Group", y = "Depression score")
```

# Cohen's d
```{r}
akp.effect(a[[2]],a[[5]],tr=0)
```

# Trimmed mean version of Cohen's d

Method from Algina et al. (2005).
```{r}
akp.effect(a[[2]],a[[5]],tr=0.2)
```

# References

Algina, J., Keselman, H. J., & Penfield, R. D. (2005). An Alternative to Cohen’s Standardized Mean Difference Effect Size: A Robust Parameter and Confidence Interval in the Two Independent Groups Case. Psychological Methods, 10(3), 317–328. https://doi.org/10.1037/1082-989X.10.3.317

Clark, F., Jackson, J., Carlson, M., Chou, C.-P., Cherry, B. J., Jordan-Marsh, M., Knight, B. G., Mandel, D., Blanchard, J., Granger, D. A., Wilcox, R. R., Lai, M. Y., White, B., Hay, J., Lam, C., Marterella, A., & Azen, S. P. (2012). Effectiveness of a lifestyle intervention in promoting the well-being of independently living older people: Results of the Well Elderly 2 Randomised Controlled Trial. J Epidemiol Community Health, 66(9), 782–790. https://doi.org/10.1136/jech.2009.099754


