---
title: "Group comparison example"
author: "Guillaume A. Rousselet & Rand R. Wilcox"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: no
    number_sections: no
    toc: yes
    toc_depth: 2
    # github_document:
    # html_preview: yes
    # toc: yes
    # toc_depth: 2
---

# Dependencies

```{r}
library(tibble)
library(ggplot2)
source("./code/functions.R")
source("./code/theme_gar.txt")
```

# Load data
```{r}
A1 <- read.table(file = "./data/A1_dat.txt")
a <- fac2list(A1$CESD,A1$edugp)
```

# Illustrate results

Horizontal black = means. 
Horizontal grey = medians.  

```{r}
df <- tibble(gp = factor(c(rep("High school", length(a[[2]])), rep("Advanced degree", length(a[[5]])))),
             score = c(a[[2]],a[[5]]))

ggplot(df, aes(x=gp, y=score)) + theme_gar +
  geom_jitter(shape = 21, width = 0.1, fill = "grey") +
  stat_summary(fun = "mean", geom = "crossbar", width = 0.5, size = 0.2) +
  stat_summary(fun = "median", geom = "crossbar", width = 0.5, size = 0.2, colour = "grey")
```

# Cohen's d
```{r}
akp.effect(a[[2]],a[[5]],tr=0)
```

# Trimmed mean version of Cohen's d

Method from Algina et al. (2005).
```{r}
akp.effect(a[[2]],a[[5]],tr=0.2)
```

# Reference

Algina, J., Keselman, H. J., & Penfield, R. D. (2005). An Alternative to Cohen’s Standardized Mean Difference Effect Size: A Robust Parameter and Confidence Interval in the Two Independent Groups Case. Psychological Methods, 10(3), 317–328. https://doi.org/10.1037/1082-989X.10.3.317

