---
title: "You get the symptoms of a replication crisis even when there isn't one: considering power"
author: "Guillaume A. Rousselet"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: no
    number_sections: no
    toc: yes
    toc_depth: 2
    # github_document:
    #   html_preview: yes
    #   toc: yes
    #   toc_depth: 2
---

# Dependencies

```{r}
library(tibble)
library(ggplot2)
library(pwr)
source("./code/functions.R")
source("./code/theme_gar.txt")
# wrapper functions around code from Yuan Yan <yuan.yan@dal.ca>
# Reference:
# Yan, Yuan, and Marc G. Genton. ‘The Tukey g-and-h Distribution’. Significance 16, no. 3 (2019): 12–13. https://doi.org/10.1111/j.1740-9713.2019.01273.x.
source("./code/makeghfig.R") 
# extra packages needed to plot the g-and-h pdf:
library(LambertW)
library(gsl)
library(nleqslv)
```

# Low replication rates even in the absence of replication issues

Perspective from Greenland et al. (2016; see also Amrhein et al. 2019):  

"if the alternative is correct and the actual power of two studies is 80%, the chance that the studies will both show $P \le 0.05$ will at best be only 0.80(0.80) = 64%; furthermore, the chance that one study shows $P \le 0.05$ and the other does not (and thus will be misinterpreted as showing conflicting results) is 2(0.80)0.20 = 32% or about 1 chance in 3."

Probability that two studies will both lead to $p \le 0.05$ = 0.80 * 0.80 = 0.64.  
Conflicting results in 2 * 0.80 * 0.20 = 0.32.

## Plot replication as a function of power

Plot probability of consistent and inconsistent results as a function of power.  

```{r}
power.vec <- seq(0.05, 0.95, 0.05)
np <- length(power.vec)

df <- tibble(x = c(power.vec, power.vec),
             y = c(power.vec^2, 2*power.vec*(1-power.vec)),
             Result = factor(rep(c("Consistent", "Inconsistent"), each = np)))

p.80 <- ggplot(df, aes(x=x, y=y, colour = Result, fill = Result)) + theme_gar +
  geom_line() +
  geom_point(shape = 21) +
  coord_cartesian(xlim = c(0,1), ylim = c(0, 1)) +
  scale_x_continuous(breaks = power.vec, minor_breaks = NULL, 
                     labels = c(".05", "",".15", "",".25", "",".35", "",".45", "",".55", "",".65", "",".75", "",".85", "",".95")) +
  scale_y_continuous(breaks = seq(0, 1, 0.1)) +
  labs(x = "Power", y = "Probability") +
  scale_fill_manual(values = c("black", "grey")) +
  scale_colour_manual(values = c("black", "grey")) +
  theme(legend.position = c(.25,.8),
        panel.grid.minor.y = element_blank())
p.80
```

When deciding about consistency between experiments using statistical significance, the probability to reach the correct decision depends on power, and unless power is very high, we will often be wrong. 

In practice, the situation is probably worse, because power analyses are typically performed assuming parametric assumptions are met, so the power will be lower than expected -- see simulations in Rousselet & Wilcox (2020); Wilcox & Rousselet (2023); Rousselet, Pernet & Wilcox (2023). The next section looks at decision consistency as a function of population skewness, which can strongly affect power.   

Why consider power as low as 5%? If that seems unrealistic, a search for n=3 or n=4 in Nature and Science magazines will reveal recent experiments carried out with very small sample sizes in the biological sciences. Also, in psychology, interactions require much larger sample sizes than often used, for instance when comparing correlation coefficients (Rousselet, Pernet & Wilcox, 2023). So very low power is still a real concern.

# How skewness affects decision consistency

Use `g-and-h` distributions. See details in Rousselet & Wilcox (2020) and Yan & Genton (2019).

## Illustrate populations

```{r, warning=FALSE}
p.gh <- plot_g_pdf(gvec = seq(0, 1, 0.1), h = 0) 
p.gh
```

## Simulation
```{r, eval=FALSE}
set.seed(777)
gvec <- seq(0, 1, 0.1) # g values to consider
ng <- length(gvec)
h <- 0 # fixed h value
nsim <- 100000 # simulation iterations
n <- 20 # sample size
targ.pow <- 0.8 # target power
# es <- 0.7 # effect size
aat <- 0.05 # arbitrary alpha threshold
es <- pwr.t.test(n=n, power=targ.pow, sig.level=aat, type="one.sample", alternative="two.sided")$d
simres.fp <- array(data = NA, dim = c(nsim, ng, 3))
simres.tp <- array(data = NA, dim = c(nsim, ng, 3))

for(G in 1:ng){
  g <- gvec[G]
  # trimmed mean of g-and-h population
  tm0 <- ghmean(g=g, h=h)$mean
  tm10 <- ghtrim(tr=0.1, g=g, h=h)
  tm20 <- ghtrim(tr=0.2, g=g, h=h)
  
  for(S in 1:nsim){
    samp <- ghdist(n, g = g, h = h)
    # one-sample t-test on trimmed means 
    # change null.value depending on amount of trimming
    # FALSE POSITIVES
    simres.fp[S,G,1] <- trimci.pval(samp, tr = 0, null.value = tm0) <= aat
    simres.fp[S,G,2] <- trimci.pval(samp, tr = 0.1, null.value = tm10) <= aat
    simres.fp[S,G,3] <- trimci.pval(samp, tr = 0.2, null.value = tm20) <= aat
    # TRUE POSITIVES
    simres.tp[S,G,1] <- trimci.pval(samp + es, tr = 0, null.value = tm0) <= aat
    simres.tp[S,G,2] <- trimci.pval(samp + es, tr = 0.1, null.value = tm10) <= aat
    simres.tp[S,G,3] <- trimci.pval(samp + es, tr = 0.2, null.value = tm20) <= aat
  }
}

save(simres.fp, simres.tp, gvec, ng, aat,
     file = "./data/nocrisis_gh.RData")
```

## Results

### Load simulation results
```{r}
load(file = "./data/nocrisis_gh.RData")
```

### False positives
```{r}
df <- tibble(g = rep(gvec, 3),
             Prob = as.vector(apply(simres.fp, c(2,3), mean)),
             Inference = factor(rep(c("0% TM", "10% TM", "20% TM"), each = ng)))

g.lab <- c("0", ".1", ".2", ".3", ".4", ".5", ".6", ".7", ".8", ".9", "1")

p.fp <- ggplot(df, aes(x=g, y=Prob, colour = Inference, fill = Inference)) + theme_gar +
  geom_hline(yintercept = aat, linetype = "dashed") +
  geom_line() +
  geom_point(shape = 21) +
  coord_cartesian(xlim = c(0,1), ylim = c(0, .2)) +
  scale_x_continuous(breaks = gvec, minor_breaks = NULL, labels = g.lab) +
  # labels = c("0", "",".1", "",".2", "",".3", "",".4", "",".5", "",".6", "",".7", "",".8", "",".9", "", "1")
  labs(x = "g value", y = "False positives") +
  scale_fill_manual(values = c("black", "darkgrey","lightgrey")) +
  scale_colour_manual(values = c("black", "darkgrey", "lightgrey")) +
  theme(legend.position = c(.2,.8))
p.fp
```

### Power
```{r}
df <- tibble(g = rep(gvec, 3),
             Prob = as.vector(apply(simres.tp, c(2,3), mean)),
             Inference = factor(rep(c("0% TM", "10% TM", "20% TM"), each = ng)))

p.tp <- ggplot(df, aes(x=g, y=Prob, colour = Inference, fill = Inference)) + theme_gar +
  geom_hline(yintercept = .8, linetype = "dashed") +
  geom_line() +
  geom_point(shape = 21) +
  coord_cartesian(xlim = c(0,1), ylim = c(0, 1)) +
  scale_x_continuous(breaks = gvec, minor_breaks = NULL, labels = g.lab) +
  # labels = c("0", "",".1", "",".2", "",".3", "",".4", "",".5", "",".6", "",".7", "",".8", "",".9", "", "1")
  scale_y_continuous(breaks = seq(0,1,0.2)) +
  labs(x = "g value", y = "True positives") +
  scale_fill_manual(values = c("black", "darkgrey","lightgrey")) +
  scale_colour_manual(values = c("black", "darkgrey", "lightgrey")) +
  theme(legend.position = c(.25,.25))
p.tp
```

### P(consistent pos)

Probability of a positive outcome in both experiments.

```{r}
df <- tibble(g = rep(gvec, 3),
             Prob = as.vector(apply(simres.tp, c(2,3), mean))^2,
             Inference = factor(rep(c("0% TM", "10% TM", "20% TM"), each = ng)))

p.cons <- ggplot(df, aes(x=g, y=Prob, colour = Inference, fill = Inference)) + 
  theme_gar +
  geom_line() +
  geom_point(shape = 21) +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
  scale_x_continuous(breaks = gvec, minor_breaks = NULL, labels = g.lab) +
  # labels = c("0", "",".1", "",".2", "",".3", "",".4", "",".5", "",".6", "",".7", "",".8", "",".9", "", "1")
  scale_y_continuous(breaks = seq(0, 1, 0.1)) +
  labs(x = "g value", y = "P(consistent pos)") +
  scale_fill_manual(values = c("black", "darkgrey","lightgrey")) +
  scale_colour_manual(values = c("black", "darkgrey", "lightgrey")) +
  theme(legend.position = c(.2,.2))
p.cons
```

### P(inconsistent)

Probability of a positive outcome in one experiment and a negative outcome in the other one.

```{r}
powres <- as.vector(apply(simres.tp, c(2,3), mean))

df <- tibble(g = rep(gvec, 3),
             Prob = 2*powres*(1-powres),
             Inference = factor(rep(c("0% TM", "10% TM", "20% TM"), each = ng)))

p.incons <- ggplot(df, aes(x=g, y=Prob, colour = Inference, fill = Inference)) + theme_gar +
  geom_line() +
  geom_point(shape = 21) +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
  scale_x_continuous(breaks = gvec, minor_breaks = NULL, labels = g.lab) +
  # labels = c("0", "",".1", "",".2", "",".3", "",".4", "",".5", "",".6", "",".7", "",".8", "",".9", "", "1")
  scale_y_continuous(breaks = seq(0, 1, 0.1)) +
  labs(x = "g value", y = "P(inconsistent)") +
  scale_fill_manual(values = c("black", "darkgrey","lightgrey")) +
  scale_colour_manual(values = c("black", "darkgrey", "lightgrey")) +
  theme(legend.position = c(.2,.8))
p.incons
```

# Summary figures

```{r, eval=FALSE}
cowplot::plot_grid(p.80 + theme(legend.position = c(.3,.8)), 
                   p.gh + theme(legend.position = c(.8, .65)) +
                     guides(colour = guide_legend(ncol = 2, byrow = TRUE,
                                                  override.aes = list(linewidth = 4))) +
                     theme(legend.key.width = unit(1, "cm")),
                   p.tp, p.fp + theme(legend.position = "none"),
                   p.cons + theme(legend.position = "none"), 
                   p.incons + theme(legend.position = "none"),
                   labels = c("A", "B", "C", "D", "E", "F"),
                   label_size = 20,
                   ncol = 2)
ggsave(filename = "./figures/nocrisis.pdf", width = 9, height = 12)
```

# Effects of outliers on decision consistency

Now we sample from a contaminated normal population (also called a mixed normal distribution), given a sample size of $n=20$. We consider one-sample t-tests on means, 10% trimmed means and 20% trimmed means.  

## Illustrate normal and contaminated normal populations

```{r}
x <- seq(-5,5,0.01)
n <- length(x)
y.norm <- dnorm(x)
y.cnorm <- 0.9*dnorm(x) + .1*dnorm(x,0,10)

df <- tibble(x = c(x,x), 
             y = c(y.norm, y.cnorm),
             Distribution = factor(rep(c("Normal", "Mixed normal"), each = n)))

df$Distribution <- keeporder(df$Distribution) 

ggplot(df, aes(x = x, y = y, colour = Distribution, linetype = Distribution)) +
  theme_gar + 
  geom_line() +
  # formatting
  scale_colour_manual(values=c("black", "grey50")) + 
  scale_linetype_manual(values = c("solid", "dashed")) +
  theme(axis.text = element_text(size = 16),
        axis.title = element_text(size = 18),
        plot.title = element_text(size=20),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank(),
        legend.position = c(0.5, 0.15),
        legend.background = element_rect(fill=alpha(0.4))) +
  labs(x = "Values") + # y = "Density") +
  scale_x_continuous(breaks = seq(-3, 3, 1)) +
  coord_cartesian(xlim = c(-3, 3))

ggsave(filename = "./figures/cnorm.pdf", width = 9, height = 5)
```

## Simulation
```{r, eval=FALSE}
set.seed(777)
nsim <- 100000 # simulation iterations
n <- 20 # sample size
targ.pow <- 0.8 # target power
aat <- 0.05 # arbitrary alpha threshold
es <- pwr.t.test(n=n, power=targ.pow, sig.level=aat, type="one.sample", alternative="two.sided")$d
simres.fp <- array(data = NA, dim = c(nsim, 2, 3)) # nsim x norm/cnrom x m/10tm/20tm
simres.tp <- array(data = NA, dim = c(nsim, 2, 3)) # nsim x norm/cnrom x m/10tm/20tm
tm0 <- 0
tm10 <- 0
tm20 <- 0

for(S in 1:nsim){
  
  # Normal population
  samp <- rnorm(n, mean = 0, sd = 1)
  # one-sample t-test on trimmed means 
  # change null.value depending on amount of trimming
  # FALSE POSITIVES
  simres.fp[S,1,1] <- trimci.pval(samp, tr = 0, null.value = tm0) <= aat
  simres.fp[S,1,2] <- trimci.pval(samp, tr = 0.1, null.value = tm10) <= aat
  simres.fp[S,1,3] <- trimci.pval(samp, tr = 0.2, null.value = tm20) <= aat
  # TRUE POSITIVES
  simres.tp[S,1,1] <- trimci.pval(samp + es, tr = 0, null.value = tm0) <= aat
  simres.tp[S,1,2] <- trimci.pval(samp + es, tr = 0.1, null.value = tm10) <= aat
  simres.tp[S,1,3] <- trimci.pval(samp + es, tr = 0.2, null.value = tm20) <= aat
  
  # Contaminated normal population
  samp <- cnorm(n)
  # one-sample t-test on trimmed means 
  # change null.value depending on amount of trimming
  # FALSE POSITIVES
  simres.fp[S,2,1] <- trimci.pval(samp, tr = 0, null.value = tm0) <= aat
  simres.fp[S,2,2] <- trimci.pval(samp, tr = 0.1, null.value = tm10) <= aat
  simres.fp[S,2,3] <- trimci.pval(samp, tr = 0.2, null.value = tm20) <= aat
  # TRUE POSITIVES
  simres.tp[S,2,1] <- trimci.pval(samp + es, tr = 0, null.value = tm0) <= aat
  simres.tp[S,2,2] <- trimci.pval(samp + es, tr = 0.1, null.value = tm10) <= aat
  simres.tp[S,2,3] <- trimci.pval(samp + es, tr = 0.2, null.value = tm20) <= aat
}

save(simres.fp, simres.tp, aat,
     file = "./data/nocrisis_cnorm.RData")
```

## Results

### Power
```{r}
load(file = "./data/nocrisis_cnorm.RData")
powres <- apply(simres.tp, c(2,3), mean)
powres
```

### Consistency for positive results
```{r}
powres^2
```

Under normality, as before given 80% power, assuming a true effect, two studies will report consistently positive results in only `r round(100*powres[1,1]^2, digits = 1)`% of the time, dropping to `r round(100*powres[1,2]^2, digits = 1)`% for 10% trimmed means, and `r round(100*powres[1,3]^2, digits = 1)`% for 20% trimmed means. 

When sampling from a contaminated normal population, now power is only `r round(100*powres[2,1], digits = 1)`% for the mean, but remains high at `r round(100*powres[2,2], digits = 1)`% for the 10% trimmed mean and `r round(100*powres[2,3], digits = 1)`% for the 20% trimmed mean. As a result, positive decision consistency drops to `r round(100*powres[2,1]^2, digits = 1)`% for the mean, `r round(100*powres[2,2]^2, digits = 1)`% for 10% trimmed means, and `r round(100*powres[2,3]^2, digits = 1)`% for 20% trimmed means.

# How skewness and outliers affect decision consistency

Use g-and-h distributions. See details in Rousselet & Wilcox (2020).
Now h = 0.1, so outliers are more likely than in the previous simulation. 

## Illustrate population
```{r}
plot_h_pdf(hvec = c(0, 0.1))
```

## Simulation
```{r, eval=FALSE}
set.seed(777)
gvec <- seq(0, 1, 0.1) # g values to consider
ng <- length(gvec)
h <- 0.1 # fixed h value
nsim <- 100000 # simulation iterations
n <- 20 # sample size
targ.pow <- 0.8 # target power
# es <- 0.7 # effect size
aat <- 0.05 # arbitrary alpha threshold
es <- pwr.t.test(n=n, power=targ.pow, sig.level=aat, type="one.sample", alternative="two.sided")$d
simres.fp <- array(data = NA, dim = c(nsim, ng, 3))
simres.tp <- array(data = NA, dim = c(nsim, ng, 3))

for(G in 1:ng){
  g <- gvec[G]
  # trimmed mean of g-and-h population
  tm0 <- ghmean(g=g, h=h)$mean
  tm10 <- ghtrim(tr=0.1, g=g, h=h)
  tm20 <- ghtrim(tr=0.2, g=g, h=h)
  
  for(S in 1:nsim){
    samp <- ghdist(n,g = g, h = h)
    # one-sample t-test on trimmed means 
    # change null.value depending on amount of trimming
    # FALSE POSITIVES
    simres.fp[S,G,1] <- trimci.pval(samp, tr = 0, null.value = tm0) <= aat
    simres.fp[S,G,2] <- trimci.pval(samp, tr = 0.1, null.value = tm10) <= aat
    simres.fp[S,G,3] <- trimci.pval(samp, tr = 0.2, null.value = tm20) <= aat
    # TRUE POSITIVES
    simres.tp[S,G,1] <- trimci.pval(samp + es, tr = 0, null.value = tm0) <= aat
    simres.tp[S,G,2] <- trimci.pval(samp + es, tr = 0.1, null.value = tm10) <= aat
    simres.tp[S,G,3] <- trimci.pval(samp + es, tr = 0.2, null.value = tm20) <= aat
  }
}

save(simres.fp, simres.tp, gvec, ng, aat,
     file = "./data/nocrisis_gh01.RData")
```

## Results

### Load simulation results
```{r}
load(file = "./data/nocrisis_gh01.RData")
```

### False positives
```{r}
df <- tibble(g = rep(gvec, 3),
             Prob = as.vector(apply(simres.fp, c(2,3), mean)),
             Inference = factor(rep(c("0% TM", "10% TM", "20% TM"), each = ng)))

ggplot(df, aes(x=g, y=Prob, colour = Inference, fill = Inference)) + theme_gar +
  geom_hline(yintercept = aat, linetype = "dashed") +
  geom_line() +
  geom_point(shape = 21) +
  coord_cartesian(xlim = c(0,1), ylim = c(0, .2)) +
  scale_x_continuous(breaks = gvec, minor_breaks = NULL) +
  # labels = c("0", "",".1", "",".2", "",".3", "",".4", "",".5", "",".6", "",".7", "",".8", "",".9", "", "1")
  labs(x = "g value", y = "False positives") +
  scale_fill_manual(values = c("black", "darkgrey","lightgrey")) +
  scale_colour_manual(values = c("black", "darkgrey", "lightgrey")) +
  theme(legend.position = c(.2,.8))
```

### Power
```{r}
df <- tibble(g = rep(gvec, 3),
             Prob = as.vector(apply(simres.tp, c(2,3), mean)),
             Inference = factor(rep(c("0% TM", "10% TM", "20% TM"), each = ng)))

ggplot(df, aes(x=g, y=Prob, colour = Inference, fill = Inference)) + theme_gar +
  geom_hline(yintercept = .8, linetype = "dashed") +
  geom_line() +
  geom_point(shape = 21) +
  coord_cartesian(xlim = c(0,1), ylim = c(0, 1)) +
  scale_x_continuous(breaks = gvec, minor_breaks = NULL) +
  # labels = c("0", "",".1", "",".2", "",".3", "",".4", "",".5", "",".6", "",".7", "",".8", "",".9", "", "1")
  scale_y_continuous(breaks = seq(0,1,0.2)) +
  labs(x = "g value", y = "True positives") +
  scale_fill_manual(values = c("black", "darkgrey","lightgrey")) +
  scale_colour_manual(values = c("black", "darkgrey", "lightgrey")) +
  theme(legend.position = c(.2,.2))
```

### P(consistent pos)
```{r}
df <- tibble(g = rep(gvec, 3),
             Prob = as.vector(apply(simres.tp, c(2,3), mean))^2,
             Inference = factor(rep(c("0% TM", "10% TM", "20% TM"), each = ng)))

ggplot(df, aes(x=g, y=Prob, colour = Inference, fill = Inference)) + theme_gar +
  geom_line() +
  geom_point(shape = 21) +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
  scale_x_continuous(breaks = gvec, minor_breaks = NULL) +
  # labels = c("0", "",".1", "",".2", "",".3", "",".4", "",".5", "",".6", "",".7", "",".8", "",".9", "", "1")
  scale_y_continuous(breaks = seq(0, 1, 0.1)) +
  labs(x = "g value", y = "P(consistent pos)") +
  scale_fill_manual(values = c("black", "darkgrey","lightgrey")) +
  scale_colour_manual(values = c("black", "darkgrey", "lightgrey")) +
  theme(legend.position = c(.2,.2))
```

### P(inconsistent)
```{r}
powres <- as.vector(apply(simres.tp, c(2,3), mean))

df <- tibble(g = rep(gvec, 3),
             Prob = 2*powres*(1-powres),
             Inference = factor(rep(c("0% TM", "10% TM", "20% TM"), each = ng)))

ggplot(df, aes(x=g, y=Prob, colour = Inference, fill = Inference)) + theme_gar +
  geom_line() +
  geom_point(shape = 21) +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
  scale_x_continuous(breaks = gvec, minor_breaks = NULL) +
  # labels = c("0", "",".1", "",".2", "",".3", "",".4", "",".5", "",".6", "",".7", "",".8", "",".9", "", "1")
  scale_y_continuous(breaks = seq(0, 1, 0.1)) +
  labs(x = "g value", y = "P(inconsistent)") +
  scale_fill_manual(values = c("black", "darkgrey","lightgrey")) +
  scale_colour_manual(values = c("black", "darkgrey", "lightgrey")) +
  theme(legend.position = c(.2,.8))
```

# References

Amrhein, V., Trafimow, D., & Greenland, S. (2019). Inferential Statistics as Descriptive Statistics: There Is No Replication Crisis if We Don’t Expect Replication. The American Statistician, 73(sup1), 262–270. https://doi.org/10.1080/00031305.2018.1543137

Greenland, S., Senn, S. J., Rothman, K. J., Carlin, J. B., Poole, C., Goodman, S. N., & Altman, D. G. (2016). Statistical tests, P values, confidence intervals, and power: A guide to misinterpretations. European Journal of Epidemiology, 31(4), 337–350. https://doi.org/10.1007/s10654-016-0149-3

Rousselet, G., Pernet, C. R., & Wilcox, R. R. (2023). An introduction to the bootstrap: A versatile method to make inferences by using data-driven simulations. Meta-Psychology, 7. https://doi.org/10.15626/MP.2019.2058

Rousselet, G. A., & Wilcox, R. R. (2020). Reaction Times and other Skewed Distributions: Problems with the Mean and the Median. Meta-Psychology, 4. https://doi.org/10.15626/MP.2019.1630

Wilcox, R. R., & Rousselet, G. A. (2023). An Updated Guide to Robust Statistical Methods in Neuroscience. Current Protocols, 3(3), e719. https://doi.org/10.1002/cpz1.719

Yan, Y., & Genton, M. G. (2019). The Tukey g-and-h distribution. Significance, 16(3), 12–13. https://doi.org/10.1111/j.1740-9713.2019.01273.x

