---
title: "Prediction intervals"
author: "Guillaume A. Rousselet"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: no
    number_sections: no
    toc: yes
    toc_depth: 2
    # github_document:
    # html_preview: yes
    # toc: yes
    # toc_depth: 2
---

# Dependencies

```{r}
library(tibble)
library(ggplot2)
# source("./code/functions.R")
```


# Prediction intervals

```{r}
predint <- function(n1, n2, m1, sd1, alpha = 0.05){
se <- sqrt( (sd1^2 / n1) + (sd1^2 / n2))
df <- n1 - 1
tval <- qt(1-alpha/2, df)
ci[1] <- m1 - qt(1-alpha/2,df)*se
ci[2] <- m1 + qt(1-alpha/2,df)*se
ci
}
```

## Example from Spence & Stanley (2024)

https://journals.sagepub.com/doi/10.1177/25152459231217932

```{r}
n1 <- 50
n2 <- 70
m1 <- 7.21
sd1 <- 2.20
ci <- predint(n1, n2, m1, sd1)
ci
```


# Simulation: normal distribution

In each simulation, perform 2 experiments. First experiment has $n1=50$, second has $n2=70$. Consider one-sample case.

Consider these rules to indicate support for a replication:  
 
- mean in sample 2, $m_2$, is contained in prediction interval;  
- $m_2$ is contained in the confidence interval from experiment 1;  
- the confidence intervals from the two experiments overlap.  

```{r}
nsim <- 10000
simres <- matrix(NA, nrow = nsim, 3)
n1 <- 50
n2 <- 70

for(S in 1:nsim){
  samp1 <- rnorm(n1)
  ci1 <- t.test(samp1)$conf.int
  m1 <- mean(samp1)
  sd1 <- sd(samp1)
  pi <- predint(n1, n2, m1, sd1)
  simres.pi.width[S] <- pi[2] - pi[1]
  samp2 <- rnorm(n2)
  ci2 <- t.test(samp2)$conf.int
  m2 <- mean(samp2)
  # rule 1: m2 contained in prediction interval
  simres[S,1] <- m2 >= pi[1] && m2 <= pi[2]
  # rule 2: m2 contained in confidence interval
  simres[S,2] <- m2 >= ci1[1] && m2 <= ci1[2]
  # rule 3: confidence intervals overlap
  simres[S,3] <- ci1[1] <= ci2[2] && ci2[1] <= ci1[2]
}
# coverage
apply(simres, 2, mean)
```

# Simulation: lognormal distribution

```{r}
nsim <- 10000
simres <- matrix(NA, nrow = nsim, 3)
n1 <- 50
n2 <- 70

for(S in 1:nsim){
  samp1 <- rlnorm(n1)
  ci1 <- t.test(samp1)$conf.int
  m1 <- mean(samp1)
  sd1 <- sd(samp1)
  pi <- predint(n1, n2, m1, sd1)
  simres.pi.width[S] <- pi[2] - pi[1]
  samp2 <- rlnorm(n2)
  ci2 <- t.test(samp2)$conf.int
  m2 <- mean(samp2)
  # rule 1: m2 is contained in prediction interval
  simres[S,1] <- m2 >= pi[1] && m2 <= pi[2]
  # rule 2: m2 is contained in confidence interval
  simres[S,2] <- m2 >= ci1[1] && m2 <= ci1[2]
  # rule 3: the confidence intervals overlap
  simres[S,3] <- ci1[1] <= ci2[2] && ci2[1] <= ci1[2]
}
# coverage
apply(simres, 2, mean)
```




