---
title: "Regression example"
author: "Guillaume A. Rousselet & Rand R. Wilcox"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: no
    number_sections: no
    toc: yes
    toc_depth: 2
    # github_document:
    # html_preview: yes
    # toc: yes
    # toc_depth: 2
---

# Dependencies

```{r}
library(tibble)
library(ggplot2)
source("./code/functions.R")
source("./code/regfun.R")
source("./code/theme_gar.txt")
library(cowplot)
```

# Load data
```{r}
A1 <- read.table(file = "./data/A1_dat.txt")
d <- cbind(A1$cort1,A1$cort3)
d <- elimna(d) # remove NA
df <- tibble(x = d[,1], y = d[,2])
```

# Regression
```{r}
reg.fit <- lm(data = df, formula = y ~ x)$coefficients
```

# Correlation
```{r}
cor.fit <- cor.test(x = df[[1]], y = df[[2]])
```

Pearson's correlation = `r cor.fit$parameter` [`r cor.fit$conf.int[1]`, `r cor.fit$conf.int[2]`], *p*=`r cor.fit$p.value`.

## Eliminate univariate outliers only

Eliminate only outliers along x-axis (leverage points).

```{r}
# d=elimna(cbind(A1$cort1,A1$cort2))
id <- outpro(d[,1])$keep
plot(d[id,1], d[id,2], xlab='Cort Time 1', ylab='Cort Time 2')
abline(ols(d[id,1], d[id,2])$coef)
pcor(d[id,1], d[id,2]) # Pearson .419
```

## Robust correlation
Use skipped correlation: remove multivariate outliers defined using robust projection method, then compute Pearson correlation using remaining points. Test statistic is adjusted to take outlier removal into account.  
```{r}
scor(x = d[,1], y = d[,2])
```

# Scatterplot

```{r}
ggplot(df, aes(x=x, y=y)) + theme_gar +
  geom_point(alpha = 0.5) +
  labs(x = "Cortisol at time 1",
       y = "Cortisol at time 2") +
  geom_abline(intercept = reg.fit[1], slope = reg.fit[2])
```

# Flag multivariate outliers 

```{r}
out.res <- outpro(d)
```

# Scatterplot with colour-coded outliers

```{r}
df <- tibble(x = d[,1], 
             y = d[,2],
             outlier = factor(!(1:length(d[,1]) %in% out.res$keep)))

pA <- ggplot(df, aes(x=x, y=y, colour = outlier, shape = outlier)) + theme_gar +
  geom_point(alpha = 0.7, size = 2) +
  scale_colour_manual(values = c("black", "grey")) + 
  labs(x = "Cortisol at time 1",
       y = "Cortisol at time 2") +
  geom_abline(intercept = reg.fit[1], slope = reg.fit[2]) +
  theme(legend.position = c(.25, .8)) 
pA
```

# Scatterplot without multivariate outliers

```{r}
df.out <- tibble(x = d[out.res$keep,1],
                 y = d[out.res$keep,2])

reg.fit.out <- lm(data = df.out, formula = y ~ x)$coefficients
#cor.fit <- cor.test(x = df.out[[1]], y = df.out[[2]])

pB <- ggplot(df.out, aes(x=x, y=y)) + theme_gar +
  geom_point(alpha = 0.7) +
  labs(x = "Cortisol at time 1",
       y = "Cortisol at time 2") +
  geom_abline(intercept = reg.fit.out[1], slope = reg.fit.out[2])
pB
```
## Merge panels
```{r, eval = FALSE}
cowplot::plot_grid(pA, pB, 
                   labels = c("A", "B"),
                   label_size = 20,
                   ncol = 2)
ggsave(filename = "./figures/reg.pdf", width = 9, height = 4)
```

